---
title: "🗺️ Fundamentals of Map Design with `mapgl`"
author: "DS 421 – Cartographic Design"
date: "2025"
format:
  html:
    toc: true
    code-fold: true
    theme: minty
    css: styles.css
    self-contained: false
execute:
  eval: false  # prevent chunk execution on render
---

Welcome to your first creative exploration of **map design using R and the `mapgl` package**! In this tutorial, you'll:

-   Load spatial data using `tidycensus`
-   Create your first interactive maps with `mapgl`
-   Learn both **continuous** and **categorical** styling methods
-   Add interactivity like **tooltips** and **hover effects**

Let’s dive in!

------------------------------------------------------------------------

## 📦 Getting Started

Make sure you’ve installed the required packages:

Load libraries:

```{r}
library(tidyverse)
library(tidycensus)
library(mapgl)
library(viridisLite)
library(RColorBrewer)
library(glue)
```

------------------------------------------------------------------------

## 🧭 Load Hawaii Median House Hold Income Data

To create a meaningful map, we need **geospatial data** — data that includes both attributes (e.g. median age) and geometry (e.g. Census tract boundaries).

The `tidycensus` package allows us to pull **American Community Survey (ACS)** data directly from the U.S. Census Bureau. When you use `geometry = TRUE`, you’re telling R to return spatial features (polygons, points, or lines) alongside the statistics.

Let’s look at what we’re loading:

```{r}
hi_median_income <- get_acs(
  geography = "tract",
  variables = "B19013_001",  # median income
  state = "HI",
  year = 2023,
  geometry = TRUE
)

```

### 👀 Explore the Data in the Viewer

Click on `hi_median_income` in your **Environment tab** or run `View(hi_median_income)` to open it as a spreadsheet.

Notice: - The `estimate` column shows the median age - The `geometry` column contains spatial polygons (`MULTIPOLYGON`) - This is an **`sf` object**, which stands for *Simple Features* — this is what makes the data "mappable" in R

> ✅ If your data has a `geometry` column and `sf` class, it’s ready for cartographic visualization!

> ✅ Now lets plot an interactive map centered on Hawaii

------------------------------------------------------------------------

### Draw a blank map of Hawaii

```{r}
hi_map <- mapboxgl(mapbox_style("light"), bounds = hi_median_income)

hi_map
```

## 🎨 Continuous Styling

One of the most important choices in cartographic design is **how to represent your data visually** — especially when it comes to **color**. Should you show your data as a **smooth gradient** or in **distinct categories**? That choice sends a message.

**Continuous color scales** use gradients to show smooth variation in a numeric variable — like temperature, elevation, or median age. These are great when:

-   You want to emphasize gradual change
-   The viewer should see subtle differences between areas
-   You want the map to feel intuitive and data-rich

For example, a continuous palette of deep blue to light blue can subtly show where the median income is lower vs. higher — without imposing artificial boundaries.

When designing with continuous color, think of it like painting with light: smooth tones invite exploration, nuance, and *feel* more natural.

Here’s how we apply a continuous color scale to median income:

```{r}
continuous_scale <- interpolate_palette(
  data = hi_median_income,
  column = "estimate",
  method = "equal",
  n = 5,
  palette = plasma
)

hi_map |>
  add_fill_layer(
    id = "hi_tracts",
    source = hi_median_income,
    fill_color = continuous_scale$expression,
    fill_opacity = 0.5
  ) |>
  add_legend(
    "Median Income in Hawaii",
    values = get_legend_labels(continuous_scale, digits = 0),
    colors = get_legend_colors(continuous_scale),
    type = "continuous"
  )
```

------------------------------------------------------------------------

### 🧩 **Challenge #1** – Try a Different Color Scheme

Replace `plasma` with another palette (e.g., `viridis`, `magma`, or `cividis`). Which one feels the most readable? Why?

I **think Viridis is the best option for color palette due to the dark to lighter gradient it naturally has, with this color palette you can easily see the higher and lower median incomes based on the lighter or darker colors**

```{r}
palette = viridis
```

Document your reasoning in a comment above your code.

------------------------------------------------------------------------

## 🧮 Manual Styling with Interpolation

Lets see the min and max of median income

```{r}
min(hi_median_income$estimate, na.rm = TRUE)
max(hi_median_income$estimate, na.rm = TRUE)
```

What does this look like with other data viz?

```{r}
ggplot(hi_median_income, aes(x = estimate)) +
  geom_histogram(fill = "steelblue", color = "black") +
  theme_minimal()
  
```

```{r}
hi_map |>
  add_fill_layer(
    id = "hi_tracts",
    source = hi_median_income,
    fill_color = interpolate(
      column = "estimate",
      values = c(20000, 250000),
      stops = c("lightblue", "darkblue"),
      na_color = "lightgrey"
    ),
    fill_opacity = 0.5
  ) |>
  add_legend(
    "Median Income in Hawaii",
    values = c(20000, 250000),
    colors = c("lightblue", "darkblue")
  )
```

------------------------------------------------------------------------

## 🔢 Categorical Styling

In contrast to smooth gradients, **classified (categorical) color scales** divide your data into clear buckets. This approach is especially helpful when:

-   You want to simplify complex data
-   The audience needs clear takeaways (e.g., policy decisions, eligibility zones)
-   You’re emphasizing **differences** between areas

Instead of saying “this place is a bit older,” a categorical map might say “this tract is in the highest 20%. for median income” This clarity is powerful — but it also means you’re drawing visual boundaries that might not exist in reality.

As a cartographer, **you are making choices that influence interpretation**. Use this power carefully.

When designing categorical maps: - Choose palettes that are **visually distinct** (e.g., `RColorBrewer::brewer.pal()`) - Keep the number of categories reasonable (usually 4–7) - Match color progression to the narrative — e.g., darker = more intense

> 🎨 A good map doesn’t just show data — it **communicates purpose** through design.

Here’s how we apply quantile classification to Hawaii's Median Income:

```{r}
q_class <- step_equal_interval(
  data = hi_median_income,
  column = "estimate",
  n = 5,
  colors = RColorBrewer::brewer.pal(5, "PRGn")
)

hi_map |>
  add_fill_layer(
    id = "hi_tracts",
    source = hi_median_income,
    fill_color = q_class$expression,
    fill_opacity = 0.5
  ) |>
  add_legend(
    "Median Income in Hawaii",
    values = get_legend_labels(q_class, digits = 0, suffix = " $"),
    colors = get_legend_colors(q_class),
    type = "categorical"
  )
```

------------------------------------------------------------------------

### 🧩 **Challenge #2** – Classify Differently

Let’s apply your design instincts!

Try switching from `step_quantile()` to one of the following:

-   `step_equal_interval()` – evenly spaced numeric breaks
-   `step_jenks()` – natural breaks that minimize variance within groups

Type `?step_quantile` in your console to see definitions

How does your choice affect the message your map conveys?

💬 *What looks more balanced? Which palette aligns with your intended story?*

Write 1–2 sentences reflecting on the **cartographic message** your choices imply.

From my understanding of these functions, all three can be very useful depending on your data structure and distribution of how you want to display it on your graphs. With these you are able to group them in equal parts based on number of data points, or data range, or use step_jenks() to optimize your display by minimizing in class variance.

------------------------------------------------------------------------

## Question: What is the poverty line for a household in Hawaii?

Can we create a map that only shows above/below the poverty line

```{r}
hi_median_income <- hi_median_income |>
  mutate(
    mit_threshold = ifelse(estimate < 107689, "Below Line", "Above Line")
  )
```

```{r}
mit_colors <- c("Below Line" = "#d73027", "Above Line" = "#1a9850")

mapboxgl(bounds = hi_median_income) |>
  add_fill_layer(
    id = "hi_tracts",
    source = hi_median_income,
    fill_color = match_expr(
      column = "mit_threshold",
      values = names(mit_colors),
      stops = unname(mit_colors)
    ),
    fill_opacity = 0.6
  ) |>
  add_legend(
    "Median Income Status",
    values = names(mit_colors),
    colors = unname(mit_colors),
    type = "categorical"
  )


```

## Question: How else can we think about the poverty line and map?

Hints : <https://livingwage.mit.edu/states/15>, <https://unitedforalice.org/hawaii>

### 🧩 **Challenge #3** – Classify Differently

Make a binary map, two colors, representing above and below a threshold of median income that you determine based on an outside source like the ones hinted above

The map above this challenge was the variation we did in class with mit as the source for the poverty line, the following changes i made sources the poverty line from united for Alice with 2 adults and 2 children

```{r}
hi_median_income <- hi_median_income |>
  mutate(
    alice_threshold = ifelse(estimate < 97608, "Below Line", "Above Line")
  )
```

```{r}
alice_colors <- c("Below Line" = "#d73027", "Above Line" = "#1a9850")

mapboxgl(bounds = hi_median_income) |>
  add_fill_layer(
    id = "hi_tracts",
    source = hi_median_income,
    fill_color = match_expr(
      column = "alice_threshold",
      values = names(alice_colors),
      stops = unname(alice_colors)
    ),
    fill_opacity = 0.6
  ) |>
  add_legend(
    "Median Income Status",
    values = names(alice_colors),
    colors = unname(alice_colors),
    type = "categorical"
  )


```

## ✨ Add Interactivity

```{r}
hi_median_income$popup <- glue("<strong>GEOID: </strong>{hi_median_income$GEOID}<br><strong>Median Income: </strong>{hi_median_income$estimate}")

hi_map |>
  add_fill_layer(
    id = "hi_tracts",
    source = hi_median_income,
    fill_color = interpolate(
      column = "estimate",
      values = c(20000, 250000),
      stops = c("lightblue", "darkblue"),
      na_color = "lightgrey"
    ),
    fill_opacity = 0.5,
    popup = "popup",
    tooltip = "estimate",
    hover_options = list(
      fill_color = "yellow",
      fill_opacity = 1
    )
  ) |>
  add_legend(
    "Median Income in Hawaii",
    values = c(20000, 250000),
    colors = c("lightblue", "darkblue")
  )
```

------------------------------------------------------------------------

### 🧩 **Challenge #4** – Customize a Tooltip

Make your map more personal and informative!

Instead of just `estimate`, add other descriptive info in your popup. For example, use `NAME` (tract name) or `GEOID` along with the estimate.

```{r}
hi_median_income$popup <- glue(
  "<strong>Tract:</strong> {hi_median_income$NAME}<br>
   <strong>GEOID:</strong> {hi_median_income$GEOID}<br>
   <strong>Median income:</strong> ${hi_median_income$estimate}"
)

hi_median_income$tooltip <- glue(
  "Tract: {hi_median_income$NAME}\nIncome: ${hi_median_income$estimate}"
)

hi_map |>
  add_fill_layer(
    id = "hi_tracts",
    source = hi_median_income,
    fill_color = interpolate(
      column = "estimate",
      values = c(20000, 250000),
      stops = c("lightblue", "darkblue"),
      na_color = "lightgrey"
    ),
    fill_opacity = 0.5,
    popup = "popup",
    tooltip = "tooltip",
    hover_options = list(
      fill_color = "blue",
      fill_opacity = 1
    )
  ) |>
  add_legend(
    "Median Income in Hawaii",
    values = c(20000, 250000),
    colors = c("lightblue", "darkblue")
  )
```

Bonus: Try replacing `fill_color = "yellow"` with a color of your choice that aligns better with your map’s aesthetic.

> 💡 Cartographic design is about more than color — it’s about **clarity, context, and care** in visual choices.

------------------------------------------------------------------------

## 🌟 Challenge #5 – Your Own Data!

Load median income, rent, or internet access for a different state using `tidycensus`. Here are a few suggestions: Use whatever state and variable you would like. You can also type this prompt into chat gpt: "what is the census code using tidycensus in R for median age?". It will most likely give you the correct variable code

| Variable   | Description              |
|------------|--------------------------|
| B01002_001 | Median age               |
| B25077_001 | Median home value        |
| B28002_004 | Households with internet |

Washington Challenge

```{r}
wa_internet <- get_acs(
  geography = "tract",
  variables = "B28002_004",
  state = "WA",
  year = 2023,
  geometry = TRUE
)

# Rename column for clarity
wa_internet <- wa_internet |>
  rename(broadband = estimate)

# Create popup and tooltip text
wa_internet$popup <- glue(
  "<strong>Tract:</strong> {wa_internet$NAME}<br>
   <strong>Broadband Access:</strong> {wa_internet$broadband} households"
)

wa_internet$tooltip <- glue(
  "Tract: {wa_internet$NAME}\nBroadband: {wa_internet$broadband}"
)

# Create a color scale
wa_scale <- interpolate_palette(
  data = wa_internet,
  column = "broadband",
  method = "equal",
  n = 4,
  palette = viridis
)

# Create the map
mapboxgl(bounds = wa_internet) |>
  add_fill_layer(
    id = "wa_internet",
    source = wa_internet,
    fill_color = wa_scale$expression,
    fill_opacity = 0.5,
    popup = "popup",
    tooltip = "tooltip",
    hover_options = list(
      fill_color = "orange",
      fill_opacity = 1
    )
  ) |>
  add_legend(
    "Households with Broadband (WA)",
    values = get_legend_labels(wa_scale),
    colors = get_legend_colors(wa_scale),
    type = "continuous"
  )
```

#### California Example

```{r}
# Example: Median income in California
ca_income <- get_acs(
  geography = "tract",
  variables = "B19013_001",
  state = "CA",
  year = 2023,
  geometry = TRUE
)

ca_scale <- interpolate_palette(
  data = ca_income,
  column = "estimate",
  method = "equal",
  n = 4,
  palette = viridisLite::viridis
)

mapboxgl(bounds = ca_income) |>
  add_fill_layer(
    id = "ca_tracts",
    source = ca_income,
    fill_color = ca_scale$expression,
    fill_opacity = 0.5
  ) |>
  add_legend(
    "Median income in California",
    values = get_legend_labels(ca_scale, digits = 0, prefix = "$"),
    colors = get_legend_colors(ca_scale),
    type = "continuous"
  )

```

🔍 Try: - Changing the variable to median home value- Switching the state (e.g., "HI", "NY", "TX") - Using a different styling function

------------------------------------------------------------------------

## ✅ Summary

You’ve now:

-   Built your first interactive map with `mapgl`
-   Explored **color palettes**, **classifications**, and **interactivity**
-   Experimented with **continuous vs categorical** styling
-   Learned how to **guide your reader** with popups and hover design

🎯 You’re on your way to designing thoughtful, expressive cartographic products.

Save and push your.qmd file to github to earn credit for the challenges!
